---
title: "Bradley-Terry Scalable Model Simulation"
output:
  pdf_document: default
  html_document: default
  html_notebook: default
---

### R Environment Setup

This section installs and loads the necessary packages for the simulation.

```{r setup, message=FALSE, warning=FALSE}
# Installing from GitHub
# install.packages("devtools") # if required
devtools::install_github("EllaKaye/BradleyTerryScalable", build_vignettes = TRUE)
library(BradleyTerryScalable)
library(Matrix)
library(dplyr)
library(ggplot2)
```


# Small-Scale Simulation
```{r small1}
data(citations)

```

### Transform the data to the btdata object the package can deal with

```{r small2}
citations_btdata <- btdata(citations)
citations_btdata
```

### Checking connectivity

```{r small3}
summary(citations_btdata)
```
### Since it is fully connected, we use a=1 in the btfit function (Maximum Likelihood Estimate)

```{r small4}
citations_fit <- btfit(citations_btdata, 1)
```


### Checking the info for the model

```{r small5}
summary(citations_fit)
```


### ---------------------Result Interpretation----------------------------------
### $call (Note: R code used to execute the model fitting)
### $item_summary (Data frame with fit results for each item)
### Note: estimate is the estimated strength parameter on the log scale. 
### Note: Higher value = higher strength/rank
### $component_summary (Note: Data frame summarizing the status of the fitted components)
### Note: num_items is the count of items in that component
### Note: iters = iterations for convergence
### ----------------------------------------------------------------------------


### Now we do a simulation using the fitted model, 100 times, and each is independent

```{r small6}
citations_sim <- simulate(citations_fit, nsim = 100, seed = 1)
```

### Display some results from the simulation (first two simulated wins matrices)
```{r small7}
citations_sim[1:2]
```
# Large-Scale Tournament Simulation
### IMPORTANT NOTE: They do not have real tournament data; they build a synthetic dataset that mimics real tournament data characteristics.

```{r big1}
set.seed(1989)
```

### Setting number of players (items)
```{r big2}
n_items <- 1000
```

### 1. Generating the Sparse Comparison Totals (Nmatrix)
### Create a Poisson distribution with lambda=1 to simulate the total number of comparisons between 1000 players.
### The formula calculates the number of unique pairs: n * (n - 1) / 2
```{r big3}
Nvalues <- rpois(n = n_items * (n_items - 1) / 2, lambda = 1)
poisdis <- data.frame(Nvalues = Nvalues)

ggplot(poisdis, aes(x = Nvalues)) +
  geom_histogram(binwidth = 1, color = "black") +
  labs(title = "Distribution of Poisson(1) Samples",
       x = "Value", y = "Frequency")
```

### logic function for later use
```{r big4}
notzero <- Nvalues > 0
```

### build a empty matrix
```{r big5}
Nmatrix_empty <- Matrix(nrow = n_items, ncol = n_items)
```

### Identify the element which is not 0, later we want to fill into the matrix
```{r big6}
ij <- which(lower.tri(Nmatrix_empty), arr.ind = TRUE)[notzero, ]
```

### fill in the matrix with the generated data
```{r big7}
Nmatrix <- sparseMatrix(
  i = ij[, 1],
  j = ij[, 2],
  x = Nvalues[notzero],
  symmetric = TRUE,
  dims = c(n_items, n_items)
)
```

### now for the BT part
### generating "hidden ability value" for each player, assuming it is normally distributed
### and we divide by 4 to reduce the sd of generated data

```{r big8}
pi_vec <- exp(rnorm(n_items) / 4)
```

### now we normalize pi_vec so its mean is 1,eliminate parameter redundancy caused by the Bradley-Terry model's scale invariance
```{r big9}
pi_vec <- pi_vec / mean(pi_vec)
```

### juz like the small data set, we do a simulation
### given the "hidden ability value" and "the number of time they played to each other"
```{r big10}
big_matrix <- simulate_BT(pi_vec, Nmatrix, nsim = 1, seed = 1)[[1]]
```

### convert the resulting matrix into a btdata object(the form this package can deal with)
```{r big11}
big_btdata <- btdata(big_matrix)
```

### check if it is fully connected
```{r big12}
summary(big_btdata)
```

### fit the model using MLE (a=1)
```{r big13}
the_model <- btfit(big_btdata, a = 1)
```

### extract the fitted ability parameters (pi)
```{r big14}
pi_fitted <- the_model$pi$full_dataset
```

### plot
```{r big15}
plot_df <- tibble(
  x = log(pi_vec[as.numeric(names(pi_fitted))]),
  y = log(pi_fitted)
)

ggplot(plot_df, aes(x, y)) +
  geom_point(alpha = 0.5) +
  geom_abline() +
  xlab("true strength") +
  ylab("maximum likelihood estimate") +
  ggtitle("1000-player simulation from a Bradley-Terry model") +
  theme(plot.title = element_text(hjust = 0.5))
```